---
title: "Poverty Index Construction and 4Ps Program Efficiency: A Multi-Method Analysis"
subtitle: "MCA Index Construction | Logistic Classification | NB Program Analysis | Budget Allocation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

---

# Setup

## Required Packages

```{r libraries}
# --- Data I/O & Manipulation ---
library(readr)
library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(stringr)
library(data.table)
library(purrr)
library(tibble)

# --- Statistical Modelling ---
library(psych)
library(MASS)
library(car)
library(caret)
library(pROC)
library(ResourceSelection)
library(DescTools)
library(AER)
library(glmmTMB)
library(performance)
library(DHARMa)

# --- Dimensionality Reduction ---
library(FactoMineR)
library(factoextra)

# --- Missing Data ---
library(mice)

# --- Visualization ---
library(ggplot2)
library(ggthemes)
library(corrplot)
library(gridExtra)
library(DT)
```

## Data Loading

> **Repository structure**: Place all required datasets under a `data/` folder  
> at the root of this project before knitting.
>
> | File | Description | Source |
> |------|-------------|--------|
> | `FIES_expenditure_data.xlsx` | Expenditure & utilities indicators | FIES 2023 |
> | `FIES_logit_data.xlsx`       | Household socioeconomic variables  | FIES 2023 |
> | `NHTS_4PS_data.xlsx`         | 4Ps beneficiary & population counts | NHTS-PR 2021 |
> | `NEP_budget_data.xlsx`       | Agency budget allocation sub-components | NEP 2021 |
> | `logistic_data.xlsx`         | Pre-processed BLRM modeling dataset | FIES 2023 (derived) |

```{r data-loading}
# OBJECTIVE 1 & 2 inputs (FIES 2023)
expenditure_data <- read_excel("data/FIES_expenditure_data.xlsx")
household_data   <- read_excel("data/FIES_logit_data.xlsx")

# OBJECTIVE 2 pre-processed dataset (household-level, BLRM-ready)
df <- read_excel("data/logistic_data.xlsx")

# OBJECTIVE 3 inputs (NHTS-PR 2021)
poor_data <- read_excel("data/NHTS_4PS_data.xlsx")

# OBJECTIVE 4 inputs (NEP 2021)
budget_data <- read_excel("data/NEP_budget_data.xlsx")
```

---

# Objective 1: MCA-Based Poverty Index Construction {#obj1}

> **Inputs**: 23 binary utilities/assets variables from FIES 2023  
> **Method**: Multiple Correspondence Analysis (MCA) with Benzécri correction  
> **Output**: Continuous [0, 100] poverty index per household

## Binary Variable Preparation

Transform raw expenditure columns into binary "spends/does not spend" indicators.
The resulting 23 columns reflect ownership of assets and access to services.

```{r mca-binary-prep}
mca_data <- expenditure_data %>%
  mutate(
    across(
      .cols  = starts_with("exp_"),
      .fns   = ~ if_else(is.na(.) | . == 0, 0, 1),
      .names = "spends_on_{str_remove(.col, 'exp_')}"
    )
  ) %>%
  dplyr::select(-starts_with("exp_")) %>%
  rename(
    has_a_house   = spends_on_housing,
    has_a_vehicle = spends_on_vehicle
  ) %>%
  mutate(across(everything(), as.factor))

cat("Binary MCA dataset dimensions:", nrow(mca_data), "households x", ncol(mca_data), "variables\n")
```

## MCA Eigenvalue Analysis & Dimension Selection

Apply the Benzécri retention rule (λ > 1/K) and compute corrected adjusted eigenvalues
to determine the number of dimensions to retain.

```{r mca-eigenvalues}
K             <- ncol(mca_data[, 4:26])   # 23 binary indicator variables
mca_threshold <- 1 / K                    # Benzécri threshold

# Run full MCA (retain all K dimensions for inspection)
mca_initial <- MCA(mca_data[, 4:26], ncp = K, graph = FALSE)

# Extract raw eigenvalues and variance percentages
eig_raw  <- mca_initial$eig[, "eigenvalue"]
pct_raw  <- mca_initial$eig[, "percentage of variance"]
cum_raw  <- mca_initial$eig[, "cumulative percentage of variance"]

# Benzécri-corrected adjusted eigenvalues
lambda_adj <- ((K / (K - 1))^2) * ((eig_raw - mca_threshold)^2)
lambda_adj[eig_raw <= mca_threshold] <- 0

pct_adj <- lambda_adj / sum(lambda_adj) * 100
cum_adj <- cumsum(pct_adj)

num_retained <- sum(eig_raw > mca_threshold)

eig_table <- tibble(
  Dimension  = seq_along(eig_raw),
  Lambda_Raw = round(eig_raw,    5),
  Pct_Raw    = round(pct_raw,    3),
  Lambda_Adj = round(lambda_adj, 5),
  Pct_Adj    = round(pct_adj,    3),
  Cum_Adj    = round(cum_adj,    3),
  Status     = if_else(eig_raw > mca_threshold, "Retained", "Dropped")
)

cat(sprintf(
  "Benzecri threshold (1/K = 1/%d): %.4f\nDimensions retained: %d\n",
  K, mca_threshold, num_retained
))
print(eig_table)
```

## Final MCA & Variable Contributions

```{r mca-final}
mca_final <- MCA(mca_data[, 4:26], ncp = num_retained, graph = FALSE)

# Variable contributions — focus on "Has" (Level = 1) categories on Dimension 1
var_contrib <- mca_final$var$contrib %>%
  as.data.frame() %>%
  rownames_to_column("Variable_Category") %>%
  mutate(
    Level    = str_extract(Variable_Category, "_[01]$") %>% str_remove("_"),
    Variable = str_remove(Variable_Category, "_[01]$")
  )

var_contrib %>%
  filter(Level == "1") %>%
  dplyr::select(Variable, Contribution = `Dim 1`) %>%
  mutate(
    `Share (%)` = sprintf("%.2f%%", (Contribution / sum(Contribution)) * 100)
  ) %>%
  arrange(desc(Contribution)) %>%
  print()
```

## MCA Score Extraction & Sign Alignment

Ensure Dimension 1 is positively oriented: higher score = higher welfare (lower poverty risk).

```{r mca-scores}
mca_scores_raw <- mca_final$ind$coord
mca_scores_std <- scale(mca_scores_raw)

# Align sign: Dim 1 should correlate positively with total income
sign_check <- cor(
  mca_scores_std[, 1],
  as.numeric(household_data$TOINC),
  method = "spearman",
  use    = "complete.obs"
)
cat(sprintf("Spearman r (Dim 1 vs. income): %.4f — sign: %s\n",
    sign_check, ifelse(sign_check > 0, "positive (no flip needed)", "negative (flipping)")))

mca_scores_aligned       <- mca_scores_std
mca_scores_aligned[, 1] <- sign(sign_check) * mca_scores_std[, 1]
```

## Rescaling to [0, 100] & Regional Poverty Thresholding

The continuous MCA score is rescaled to [0, 100]. Within each region, the bottom
percentile corresponding to the official PSA 2023 poverty incidence rate is classified
as poor (`mca_binary = 1`).

```{r mca-index-rescale}
rescale_0_100    <- function(x) ((x - min(x)) / (max(x) - min(x))) * 100
mca_index_dim1   <- rescale_0_100(mca_scores_aligned[, 1])
household_data$mca_index <- mca_index_dim1

# Descriptive check by income quintile — index should increase with quintile
household_data %>%
  group_by(income_q) %>%
  summarise(
    Median_MCA = round(median(mca_index, na.rm = TRUE), 2),
    Mean_MCA   = round(mean(mca_index,   na.rm = TRUE), 2),
    N          = n(),
    .groups    = "drop"
  ) %>%
  print()
```

```{r poverty-thresholding}
# Official PSA 2023 regional poverty incidence rates
region_rates <- tibble::tribble(
  ~W_REGN, ~Region_Name,    ~Poverty_Rate_2023,
  13, "NCR",          1.1,
  14, "CAR",          4.4,
   1, "Region I",     8.4,
   2, "Region II",    7.2,
   3, "Region III",   5.7,
   4, "Region IVA",   5.3,
  17, "MIMAROPA",    16.3,
   5, "Region V",    20.3,
   6, "Region VI",    9.8,
  18, "NIR",         22.6,
   7, "Region VII",  12.3,
   8, "Region VIII", 20.3,
   9, "Region IX",   24.2,
  10, "Region X",    18.4,
  11, "Region XI",   11.3,
  12, "Region XII",  17.0,
  16, "Caraga",      14.9,
  19, "BARMM",       23.5
)

logistic_data <- household_data %>%
  mutate(W_REGN = as.numeric(W_REGN)) %>%
  left_join(region_rates %>% mutate(W_REGN = as.numeric(W_REGN)), by = "W_REGN") %>%
  group_by(W_REGN) %>%
  mutate(
    MCA_threshold = quantile(mca_index,
                             probs  = first(Poverty_Rate_2023) / 100,
                             na.rm  = TRUE),
    mca_binary    = ifelse(mca_index <= MCA_threshold, 1, 0)
  ) %>%
  ungroup()

# Verification: computed incidence rates should match official PSA figures
verification_check <- logistic_data %>%
  group_by(Region_Name, Poverty_Rate_2023) %>%
  summarise(
    Computed_Incidence = round(mean(mca_binary, na.rm = TRUE) * 100, 2),
    .groups = "drop"
  )
print(verification_check)
```

---

# Objective 2: Binary Logistic Regression Model (BLRM) {#obj2}

> **Inputs**: 13 household-level indicator variables from FIES 2023  
> **Response**: `poor_MCA` (1 = poor, 0 = not poor; derived from MCA index threshold)  
> **Method**: GLM logistic regression with Recursive Feature Elimination (RFE) for variable selection

## Data Preparation & Type Coercion

```{r blrm-data-prep}
numeric_cols <- c(
  "WAGES", "TOTEX", "Male_Count", "Female_Count",
  "FSIZE", "TOINC", "URB", "Married_Count",
  "At_Least_HS_Grad_Count", "OFW_Worker_Count",
  "Has_Job_Or_Work_Count", "Looking_For_Add_Work_Count"
)

df[numeric_cols] <- lapply(df[numeric_cols], as.numeric)
df$poor_MCA      <- as.factor(df$mca_binary)

# Top-coding sparse count variables to prevent complete separation
df$OFW_Capped  <- ifelse(df$OFW_Worker_Count          >= 2, 2, df$OFW_Worker_Count)
df$Job_Capped  <- ifelse(df$Has_Job_Or_Work_Count      >= 9, 9, df$Has_Job_Or_Work_Count)
df$Look_Capped <- ifelse(df$Looking_For_Add_Work_Count >= 5, 5, df$Looking_For_Add_Work_Count)

cat("Response distribution:\n")
print(table(df$poor_MCA))
```

## Exploratory Summary Statistics

```{r blrm-eda}
eda_cols <- c(
  "WAGES", "TOTEX", "Male_Count", "Female_Count",
  "FSIZE", "TOINC", "URB", "Married_Count",
  "At_Least_HS_Grad_Count", "OFW_Capped", "Job_Capped", "Look_Capped"
)

eda_summary <- df %>%
  dplyr::select(all_of(eda_cols)) %>%
  pivot_longer(everything(), names_to = "Predictor", values_to = "Value") %>%
  group_by(Predictor) %>%
  summarise(
    Mean   = round(mean(Value,   na.rm = TRUE), 4),
    Median = round(median(Value, na.rm = TRUE), 4),
    SD     = round(sd(Value,     na.rm = TRUE), 4),
    Min    = round(min(Value,    na.rm = TRUE), 4),
    Max    = round(max(Value,    na.rm = TRUE), 4),
    .groups = "drop"
  )

print(eda_summary)
```

## Initial Full Model & Multicollinearity Check (VIF)

```{r blrm-initial-model}
logit_initial <- glm(
  poor_MCA ~ FSIZE + TOINC + URB + Married_Count +
    At_Least_HS_Grad_Count + OFW_Capped + Job_Capped + Look_Capped,
  data   = df,
  family = binomial(link = "logit")
)

cat("VIF — Initial Full Model:\n")
print(round(vif(logit_initial), 4))
```

## Variable Selection via Recursive Feature Elimination (RFE)

5-fold cross-validated RFE using GLM (logistic) and AUC as the selection criterion.

```{r blrm-rfe, cache=TRUE}
predictors <- c(
  "WAGES", "TOTEX", "Male_Count", "Female_Count",
  "FSIZE", "TOINC", "URB", "Married_Count",
  "At_Least_HS_Grad_Count", "OFW_Capped", "Job_Capped", "Look_Capped"
)

df_ml              <- df
df_ml$poor_MCA_Label <- factor(
  ifelse(df_ml$poor_MCA == "1", "Poor", "Not_Poor"),
  levels = c("Not_Poor", "Poor")
)

glm_control <- rfeControl(
  functions = caretFuncs,
  method    = "cv",
  number    = 5,
  verbose   = FALSE
)

set.seed(2026)
rfe_results <- rfe(
  x          = df_ml[, predictors],
  y          = df_ml$poor_MCA_Label,
  sizes      = seq_len(length(predictors)),
  rfeControl = glm_control,
  method     = "glm",
  family     = "binomial",
  metric     = "ROC",
  trControl  = trainControl(summaryFunction = twoClassSummary, classProbs = TRUE)
)

selected_vars <- predictors(rfe_results)
cat("Selected predictors after RFE:\n")
print(selected_vars)
```

## Final Logistic Regression Model

```{r blrm-final-model}
final_formula <- as.formula(paste("poor_MCA ~", paste(selected_vars, collapse = " + ")))

final_logit <- glm(final_formula, data = df, family = binomial(link = "logit"))
summary(final_logit)

cat("\nVIF — Final Model:\n")
print(round(vif(final_logit), 4))
```

## Model Diagnostics

```{r blrm-diagnostics, fig.width=7, fig.height=5}
# McFadden's Pseudo-R²
cat("McFadden's Pseudo-R²:\n")
print(PseudoR2(final_logit, which = "McFadden"))

# Standardized residuals
std_resid     <- rstandard(final_logit)
outlier_count <- sum(abs(std_resid) > 3)
cat(sprintf("\nObservations |std. residual| > 3: %d\n", outlier_count))

# ROC Curve & AUC
predicted_probs <- predict(final_logit, type = "response")
roc_obj         <- roc(df$poor_MCA, predicted_probs)
auc_val         <- auc(roc_obj)

cat(sprintf("AUC: %.4f\n", auc_val))

plot(
  roc_obj,
  col  = "red",
  lwd  = 2,
  main = paste0("ROC Curve — Final Logistic Model (AUC = ", round(auc_val, 3), ")")
)
abline(a = 0, b = 1, lty = 2, col = "gray50")

# Confusion Matrix (threshold = 0.50)
predicted_class <- ifelse(predicted_probs > 0.5, "1", "0")
conf_matrix     <- confusionMatrix(
  as.factor(predicted_class), df$poor_MCA, positive = "1"
)
print(conf_matrix)
```

---

# Objective 3: Provincial Negative Binomial Loglinear Model {#obj3}

> **Inputs**: 2023 Poor Index (from Obj 1 & 2), 2021 Poor Index (NHTS-PR 2021),  
>   2-Year 4Ps Program Penetration, 2021 & 2023 Population (province-level)  
> **Response**: `Poor_Counts_2023` — weighted provincial poor population  
> **Method**: Negative Binomial loglinear model with log-population offset

## Aggregate 2023 MCA-Derived Poor Counts by Province

```{r obj3-aggregation}
mca_aggregated <- logistic_data %>%
  mutate(
    weighted_poor_pop = as.numeric(mca_binary) * as.numeric(MEM_RFACT),
    W_REGN            = as.numeric(W_REGN),
    W_PROV            = as.numeric(W_PROV)
  ) %>%
  group_by(W_REGN, W_PROV) %>%
  summarise(
    Poor_Counts_2023 = round(sum(weighted_poor_pop, na.rm = TRUE)),
    .groups = "drop"
  )

cat("Provincial poor counts aggregated:", nrow(mca_aggregated), "provinces\n")
```

## Build Provincial Loglinear Dataset

Merge NHTS-PR 2021 data with the MCA-derived 2023 poor counts, then compute
penetration and poverty rate covariates.

```{r obj3-data-build}
loglinear_data <- poor_data %>%
  mutate(
    W_REGN = as.numeric(W_REGN),
    W_PROV = as.numeric(W_PROV)
  ) %>%
  dplyr::select(-Poor_Counts_2023) %>%
  inner_join(mca_aggregated, by = c("W_REGN", "W_PROV")) %>%
  mutate(
    # Scaled poverty rate in 2021
    Poverty_Rate_2021    = scale(Poor_Counts_2021 / Population_Counts_2021),
    # 4Ps penetration rates
    Penetration_4Ps_2021 = Beneficiary_Counts_2021 / Poor_Counts_2021,
    Penetration_4Ps_2023 = Beneficiary_Counts_2023 / Poor_Counts_2023,
    # Scaled 2-year change in penetration
    Penetration_OverTime = scale(Penetration_4Ps_2023 - Penetration_4Ps_2021)
  ) %>%
  group_by(W_REGN, W_PROV)

cat("Provincial dataset dimensions:", nrow(loglinear_data), "x", ncol(loglinear_data), "\n")
cat("\nColumn sums — key count variables:\n")
print(colSums(loglinear_data[, 4:9]))
```

## Baseline Poisson Model

Fit a Poisson model first to test for overdispersion before switching to NB.

```{r obj3-poisson}
provincial_poisson_model <- glmmTMB(
  Poor_Counts_2023 ~ Poverty_Rate_2021 + Penetration_OverTime +
    offset(log(Population_Counts_2023)),
  data   = loglinear_data,
  family = poisson(link = "log")
)
summary(provincial_poisson_model)

provincial_null_poisson <- glmmTMB(
  Poor_Counts_2023 ~ 1 + offset(log(Population_Counts_2023)),
  data   = loglinear_data,
  family = poisson(link = "log")
)

pearson_ratio_p <- sum(residuals(provincial_poisson_model, type = "pearson")^2) /
  (nrow(loglinear_data) - length(fixef(provincial_poisson_model)$cond))

cat(sprintf("Poisson deviance      : %.3f\n", deviance(provincial_poisson_model)))
cat(sprintf("Null Poisson deviance : %.3f\n", deviance(provincial_null_poisson)))
cat(sprintf("Pearson χ²/df ratio   : %.3f  (>> 1 → overdispersion)\n", pearson_ratio_p))

check_overdispersion(provincial_poisson_model)
```

## Negative Binomial Model

```{r obj3-nb}
provincial_nb_model <- glmmTMB(
  Poor_Counts_2023 ~ Poverty_Rate_2021 + Penetration_OverTime +
    offset(log(Population_Counts_2023)),
  data   = loglinear_data,
  family = nbinom2(link = "log")
)
s_nb <- summary(provincial_nb_model)
print(s_nb)

provincial_null_nb <- glmmTMB(
  Poor_Counts_2023 ~ 1 + offset(log(Population_Counts_2023)),
  data   = loglinear_data,
  family = nbinom2(link = "log")
)

# McFadden's Pseudo-R²
mcfadden_r2 <- 1 - (as.numeric(logLik(provincial_nb_model)) /
                    as.numeric(logLik(provincial_null_nb)))
cat(sprintf("\nMcFadden Pseudo-R² (NB): %.4f\n", mcfadden_r2))
```

## Model Comparison: Poisson vs. Negative Binomial

```{r obj3-model-comparison}
print(anova(provincial_poisson_model, provincial_nb_model))
```

## Model Fit & Diagnostic Statistics

```{r obj3-diagnostics}
cat(strrep("=", 60), "\n")
cat("  DIAGNOSTIC: MODEL FIT STATISTICS\n")
cat(strrep("=", 60), "\n")

nb_df      <- nrow(loglinear_data) - length(fixef(provincial_nb_model)$cond)
nb_dev     <- deviance(provincial_nb_model)
nb_pearson <- sum(residuals(provincial_nb_model, type = "pearson")^2)
obs_pred_r <- cor(loglinear_data$Poor_Counts_2023, fitted(provincial_nb_model))

cat(sprintf("AIC                       : %.2f\n",  AIC(provincial_nb_model)))
cat(sprintf("BIC                       : %.2f\n",  BIC(provincial_nb_model)))
cat(sprintf("Residual deviance         : %.3f  (df = %d)\n", nb_dev, nb_df))
cat(sprintf("Deviance ratio (dev/df)   : %.3f  (ideal ≈ 1)\n", nb_dev / nb_df))
cat(sprintf("Pearson χ²                : %.3f\n",  nb_pearson))
cat(sprintf("Pearson χ²/df ratio       : %.3f  (ideal ≈ 1)\n", nb_pearson / nb_df))
cat(sprintf("Obs vs. Predicted r       : %.4f\n",  obs_pred_r))
cat(sprintf("Theta (dispersion param.) : %.4f\n",  s_nb$sigma))

cat("\n", strrep("=", 60), "\n")
cat("  DIAGNOSTIC: MULTICOLLINEARITY (VIF)\n")
cat(strrep("=", 60), "\n")

tryCatch({
  vif_vals <- car::vif(provincial_nb_model)
  print(round(vif_vals, 4))
  if (any(vif_vals > 5)) {
    cat("WARNING — VIF > 5 for:", names(vif_vals)[vif_vals > 5], "\n")
  } else {
    cat("OK — All VIF < 5: no concerning multicollinearity.\n")
  }
}, error = function(e) {
  cat("car::vif() unavailable — computing from model matrix correlation.\n")
  X       <- model.matrix(provincial_nb_model)
  X_pred  <- X[, -1, drop = FALSE]
  vif_man <- diag(solve(cor(X_pred)))
  print(round(vif_man, 4))
})
```

## DHARMa Simulation-Based Residuals

```{r obj3-dharma, fig.width=11, fig.height=4.5}
sim_output <- simulateResiduals(fittedModel = provincial_nb_model)

dharma_df <- data.frame(
  fitted    = sim_output$fittedPredictedResponse,
  residuals = sim_output$scaledResiduals
)

p_qq <- ggplot(dharma_df, aes(sample = residuals)) +
  stat_qq(distribution = qunif, size = 2, alpha = 0.6, color = "steelblue") +
  stat_qq_line(distribution = qunif, color = "red", linewidth = 1) +
  labs(
    title = "DHARMa Q-Q Plot (Uniform)",
    x     = "Expected",
    y     = "Observed Scaled Residuals"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_scatter <- ggplot(dharma_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6, size = 2, color = "steelblue") +
  geom_hline(yintercept = c(0.25, 0.5, 0.75),
             linetype = "dashed", color = "gray50") +
  labs(
    title = "Residuals vs. Predicted",
    x     = "Fitted Values",
    y     = "Scaled Residuals"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

grid.arrange(p_qq, p_scatter, ncol = 2)
```

---

# Objective 4: Budget Allocation Efficiency — Exploratory NB Models {#obj4}

> **Inputs**: 6 budget allocation sub-components from NEP 2021  
>   (DepEd, DOH, DOLE, DSWD) joined to regional aggregates  
> **Response**: `Poor_Counts_2023` at the regional level  
> **Method**: Agency-specific Negative Binomial loglinear models  
> ⚠ **Note**: Regional sample is small (N = 17 after BARMM exclusion).
> Treat results as exploratory.

## Regional Data Aggregation & Budget Feature Engineering

```{r obj4-data-prep}
reg_loglinear_data <- loglinear_data %>%
  filter(W_REGN != 19) %>%           # Exclude BARMM (data quality)
  group_by(W_REGN) %>%
  summarise(
    Population_Counts_2021  = round(sum(Population_Counts_2021,  na.rm = TRUE)),
    Population_Counts_2023  = round(sum(Population_Counts_2023,  na.rm = TRUE)),
    Poor_Counts_2021        = round(sum(Poor_Counts_2021,        na.rm = TRUE)),
    Poor_Counts_2023        = round(sum(Poor_Counts_2023,        na.rm = TRUE)),
    Beneficiary_Counts_2021 = round(sum(Beneficiary_Counts_2021, na.rm = TRUE)),
    Beneficiary_Counts_2023 = round(sum(Beneficiary_Counts_2023, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(W_REGN = as.numeric(W_REGN)) %>%
  left_join(
    budget_data %>% mutate(UACS_REG_ID = as.numeric(UACS_REG_ID)),
    by = c("W_REGN" = "UACS_REG_ID")
  ) %>%
  mutate(
    # Poverty & penetration covariates (unscaled at regional level)
    Poverty_Rate_2021    = Poor_Counts_2021 / Population_Counts_2021,
    Penetration_4Ps_2021 = Beneficiary_Counts_2021 / Poor_Counts_2021,
    Penetration_4Ps_2023 = Beneficiary_Counts_2023 / Poor_Counts_2023,
    Penetration_OverTime = Penetration_4Ps_2023 - Penetration_4Ps_2021,

    # DepEd: education-focused transfers
    Total_DepEd         = Col_2_DepEd_Cash_Allowance + Col_3_DepEd_Learner_Support + Col_4_DepEd_Remaining,
    log_Total_DepEd     = log(Total_DepEd),
    DepEd_Share_Cash    = Col_2_DepEd_Cash_Allowance  / Total_DepEd,
    DepEd_Share_Support = Col_3_DepEd_Learner_Support / Total_DepEd,

    # DOH: health-focused transfers
    Total_DOH           = Col_5_DOH_Local_Health + Col_6_DOH_Public_Health + Col_7_DOH_Remaining,
    log_Total_DOH       = log(Total_DOH),
    DOH_Share_Local     = Col_5_DOH_Local_Health  / Total_DOH,
    DOH_Share_Public    = Col_6_DOH_Public_Health / Total_DOH,

    # DOLE: labor-focused transfers
    Total_DOLE            = Col_8_DOLE_Job_Search + Col_9_DOLE_Workers_Org + Col_10_DOLE_Remaining,
    log_Total_DOLE        = log(Total_DOLE),
    DOLE_Share_JobSearch  = Col_8_DOLE_Job_Search  / Total_DOLE,
    DOLE_Share_WorkersOrg = Col_9_DOLE_Workers_Org / Total_DOLE,

    # DSWD: social protection transfers
    Total_DSWD          = Col_11_DSWD_Pantawid + Col_12_DSWD_NHTS + Col_13_DSWD_Remaining,
    log_Total_DSWD      = log(Total_DSWD),
    DSWD_Share_Pantawid = Col_11_DSWD_Pantawid / Total_DSWD,
    DSWD_Share_NHTS     = Col_12_DSWD_NHTS     / Total_DSWD
  )

cat("Regional dataset:", nrow(reg_loglinear_data), "regions x", ncol(reg_loglinear_data), "columns\n")
```

## DepEd Allocative Efficiency Model

```{r obj4-deped}
DEPED_model <- glmmTMB(
  Poor_Counts_2023 ~
    Poverty_Rate_2021 + Penetration_OverTime +
    log_Total_DepEd + DepEd_Share_Support +
    offset(log(Population_Counts_2023)),
  data   = reg_loglinear_data,
  family = nbinom2(link = "log")
)
summary(DEPED_model)
cat(sprintf("DepEd model deviance: %.3f\n", deviance(DEPED_model)))
```

## DOLE Allocative Efficiency Model

```{r obj4-dole}
DOLE_model <- glmmTMB(
  Poor_Counts_2023 ~
    Poverty_Rate_2021 + Penetration_OverTime +
    log_Total_DOLE + DOLE_Share_JobSearch +
    offset(log(Population_Counts_2023)),
  data   = reg_loglinear_data,
  family = nbinom2(link = "log")
)
summary(DOLE_model)
cat(sprintf("DOLE model deviance: %.3f\n", deviance(DOLE_model)))
```

## DSWD Allocative Efficiency Model

```{r obj4-dswd}
DSWD_model <- glmmTMB(
  Poor_Counts_2023 ~
    Poverty_Rate_2021 + Penetration_OverTime +
    log_Total_DSWD + DSWD_Share_Pantawid +
    offset(log(Population_Counts_2023)),
  data   = reg_loglinear_data,
  family = nbinom2(link = "log")
)
summary(DSWD_model)
cat(sprintf("DSWD model deviance: %.3f\n", deviance(DSWD_model)))
```

---

*All datasets used in this analysis are proprietary. Data files are **not** included  
in this repository. See `data/README.md` for data availability and access instructions.*
